```{R setup, include=FALSE}
munrodata <- tidytuesdayR::tt_load('2025-08-19')
metdata <- tidytuesdayR::tt_load('2025-10-21')
library(rnaturalearth)
library(sf)
library(tidyverse)
library(dplyr)
library(tmap)
library(knitr)
```

# Question #1: How many peaks currently listed as Munros have always been included on the list?

```{R munro-analysis, echo=TRUE, message=FALSE, warning=FALSE}
scottish_munros <- munrodata$scottish_munros
munro_classification <- scottish_munros %>%
    select(Name, xcoord, ycoord, "1891", "1921", "1933", "1953", "1969", "1974", "1981", "1997", "2021") %>%
    pivot_longer(cols = c("1891", "1921", "1933", "1953", "1969", "1974", "1981", "1997", "2021"), names_to = "Year", values_to = "Classification") %>%
    group_by(Name)
munro_always <- munro_classification %>%
    summarize(Always_Munro = all(Classification == "Munro"), xcoord = first(xcoord), ycoord = first(ycoord))
    kable(head(munro_always))
munro_always_sf <- drop_na(munro_always) %>% st_as_sf(coords = c("xcoord", "ycoord"), crs = 27700, remove = FALSE)
scotland <- ne_states(geounit = "Scotland", returnclass = "sf") %>%
    st_transform(crs = 27700)

tm_shape(scotland) +
    tm_polygons(col = "lightgrey") +
    tm_shape(munro_always_sf) +
    tm_dots(col = "Always_Munro", palette = c("red", "blue"), size = 0.1) +
    tm_title("Which Monros have always been classified as such?") +
    tm_layout(legend.outside = TRUE)
```

# Question 2: Which year saw the largest number of changes to the classification?

```{R reclassification-analysis, echo=TRUE, message=FALSE, warning=FALSE}
# which years saw the most reclassifications from one type to another?
reclassifications <- munro_classification %>%
    arrange(Name, Year) %>%
    group_by(Name) %>%
    mutate(Previous_Classification = lag(Classification)) %>%
    filter(!is.na(Previous_Classification) & Classification != Previous_Classification) %>%
    ungroup() %>%
    group_by(Year) %>%
    summarize(Reclassifications = n())
ggplot(reclassifications, aes(x = Year, y = Reclassifications)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "Number of Munro Reclassifications by Year",
         x = "Year",
         y = "Number of Reclassifications") +
    theme_minimal()
```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
#install.packages("tidytuesdayR")
#install.packages("devtools")
library(tidytuesdayR)
library(rnaturalearth)
library(sf)
library(tidyverse)
library(dplyr)
library(tmap)
library(rnaturalearth)
library(knitr)
```



```{r, echo=TRUE, message=FALSE, warning=FALSE}
tuesdata <- tidytuesdayR::tt_load('2025-08-19')
scottish_munros <- tuesdata$scottish_munros

munro_map <- scottish_munros %>%
  filter(!is.na(xcoord) & !is.na(ycoord)) %>%
  st_as_sf(coords=c("xcoord","ycoord"), crs = 27700) %>%
  st_transform(4326)

projection <- scottish_munros %>%
  filter(!is.na(xcoord) & !is.na(ycoord)) %>%
  st_as_sf(coords=c("xcoord","ycoord"), crs = 27700) %>%
  st_transform(4326) %>%
  st_coordinates()

munro_map_XY <- munro_map %>%
  cbind(projection)
```

# Question #4: Which are the rainiest/coldest/sunniest/hottest months to visit the highest peak? Has that changed over time

```{r, echo=TRUE, message=FALSE, warning=FALSE}
highest_peak <- scottish_munros %>%
  filter(Height_ft == max(Height_ft, na.rm = TRUE))

tuesdata <- tidytuesdayR::tt_load('2025-10-21')

historic_station_met <- tuesdata$historic_station_met
station_meta <- tuesdata$station_meta
 
highest_coords <- highest_peak %>%
  st_as_sf(coords = c("xcoord", "ycoord"), crs = 27700) %>%
  st_transform(4326)

stations_sf <- station_meta %>%
  filter(!is.na(lng) & !is.na(lat)) %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326)

nearest_station <- stations_sf[which.min(st_distance(highest_coords, stations_sf)), ]

station_id <- nearest_station$station

weather_highest <- historic_station_met %>%
  filter(station == "dunstaffnage")

monthly_climate <- weather_highest %>%
  group_by(month) %>%
  summarise(
    avg_tmin = mean(tmin, na.rm = TRUE),
    avg_tmax = mean(tmax, na.rm = TRUE),
    total_af = sum(af, na.rm = TRUE),
    total_rain = sum(rain, na.rm = TRUE),
    total_sun = sum(sun, na.rm = TRUE)
)
coldest_month <- monthly_climate %>% filter(avg_tmin == min(avg_tmin))
hottest_month <- monthly_climate %>% filter(avg_tmax == max(avg_tmax))
rainiest_month <- monthly_climate %>% filter(total_rain == max(total_rain))
sunniest_month <- monthly_climate %>% filter(total_sun == max(total_sun))
frostiest_month <- monthly_climate %>% filter(total_af == max(total_af))

weather_highest %>%
  group_by(month) %>%
  summarise(
    avg_tmin = mean(tmin, na.rm = TRUE),
    avg_tmax = mean(tmax, na.rm = TRUE),
    avg_rain = mean(rain, na.rm = TRUE),
    avg_sun = mean(sun, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("avg"), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = month, y = value, fill = variable)) +
  geom_col(position = "dodge") +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Average Monthly Climate at Dunstaffnage",
       x = "Month", y = "Value") +
  theme_minimal()
```


# Question #5: Were there any historic years that were particularl rainy/cold/sunny/hot on the peak

```{r, echo=TRUE, message=FALSE, warning=FALSE}
weather_highest %>%
  group_by(year) %>%
  summarise(
    avg_tmin = mean(tmin, na.rm = TRUE),
    avg_tmax = mean(tmax, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = avg_tmin, color = "Min Temp")) +
  geom_line(aes(y = avg_tmax, color = "Max Temp")) +
  labs(title = "Yearly Temperature Trends at Dunstaffnage",
       y = "Temperature (°C)", color = "Legend") +
  theme_minimal()

weather_highest %>%
  group_by(year) %>%
  summarise(
    total_rain = sum(rain, na.rm = TRUE),
    total_sun = sum(sun, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = total_rain, color = "Rainfall (mm)")) +
  geom_line(aes(y = total_sun, color = "Sunshine Hours")) +
  labs(title = "Annual Rainfall and Sunshine at Dunstaffnage",
       y = "Total per Year", color = "Legend") +
  theme_minimal()

weather_highest %>%
  group_by(year) %>%
  summarise(total_af = sum(af, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = total_af)) +
  geom_line(color = "steelblue") +
  labs(title = "Total Air Frost Days per Year",
       y = "Air Frost Days", x = "Year") +
  theme_minimal()
```